Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-08 numerical-mhd-ct.tex
Finished up my edits on the fluid equations.

This MHD method was developed by \citet{Collins10} in order to study Population
III stars \citep{Xu08}, molecular clouds \citep{Collins12a}, and galaxy clusters
\citep{Xu12, Skillman13}.  A full description and suite of test problems can be
found in \citet{Collins10}

The divergence of the magnetic field, \Divb, is
identically zero due to the fact that the evolution of the magnetic field is the
curl of a vector, and the divergence of the curl of a vector is identically
zero:
\begin{eqnarray}
\partial \Bvec/\partial t & =& \frac{1}{a} \curl \Evec\\
\Divb & =& 0.
\end{eqnarray}

The Constrained Transport (CT) method \citep{Evans88, Balsara99} for
magnetohydrodynamic (MHD) evolution employs this same vector property
to evolve the magnetic field in a manner that preserves $\Divb=0$ to
limits of machine precision.  The electric field is computed using
the fluxes from the Riemann solver.  The curl of that electric field
is then used to update the magnetic field.  The advantage of this
method is that it preserves \Divb\ to machine precision.  The drawback
is twofold: increased algorithm complexity, as will be described in
this section; and since only the update of the magnetic field is
divergence free, any monopoles created by other numerical sources
persist.

\enzo\ MHD-CT has been described in great detail in \citet{Collins10},
and we direct the interested reader to that reference.  The base
Godunov method, described in \citet{Li08a}, uses spatially and
temporally second order reconstruction (both MUSCL-Hancock and
Piecewise Linear Method), a selection of Riemann solvers including
HLLC and HLLD \citep{Mignone07}.  The constrainted transport methods
are the first-order method described by \citet{Balsara99} and the
second order methods described in \citet{Gardiner05}.  The AMR
machinery is described by \citet{Balsara01} and \citet{Collins10}.

@@

Cosmological MHD can be treated in a number of ways.  As described in Section
\red{reference}, the MUSCL with Dedner treats the cosmological dilution of
$\vecB$ with a source term, 
$$ 
\frac{\partial \vecB}{\partial t} - \frac{1}{a}  \curl (\vecv \times
\vecB)  =  -\frac{\dot{a}}{2 a} \vecB. \label{eq:induction_cosmo_source}
$$
However, one can eliminate the right hand side of Equation 
\ref{eq:induction_cosmo_source} by defining
$\vecB^\prime = \vecB / \sqrt{a}.$
Then Equation \ref{eq:induction_cosmo_source} becomes
$$ 
\frac{\partial \vecB^\prime}{\partial t} - \frac{1}{a}  \curl (\vecv \times
\vecB^\prime)  = 0, \label{eq:induction_cosmo_prime_nosource}
$$
and we can reuse each part of the non-cosmological algorithm (hyperbolic solver,
electric field creation, interpolation, projection, and flux correction) without
modification.

Dual energy formalism has also been incorporated in two methods, one internal
energy, similar to that described in Section \ref{sec.hydro.ppm}, and one that
uses entropy.  The additional equations are
\begin{eqnarray}
\end{eqnarray}
\begin{eqnarray}
  \frac{\partial S}{\partial t} +\frac{1}{a} \div (S \vecv) & = &  -
  \frac{ 3 (\gamma - 1) \dot{a}}{a} S \\
  \frac{\partial \rho e}{\partial t} + \frac{1}{a} \div (\rho e \vecv) &
  = & -\frac{ 3(\gamma -1 ) \dot{a}}{a} + \frac{p}{a} \div \vecv,
\end{eqnarray}
where $e$ is the internal energy and $S=p/\rho^{\gamma-1}$ is the entropy.  If
the ratio of thermal to kinetic energy, $\eta$, is below a certain value, the
alternative energy is used.  As in \citet{Li08a}, we use $\eta=0.008$.

\enzo's MHD-CT has been used in a variety of contexts, from galaxy
cluster evolution \citep{Xu11,Xu12}, synthetic observations of radio
halos \citep{Skillman13}, primordial star formation \citep{Xu08}, and
present-day star formation \citep{Collins11, Collins12a}.

>>>>

The third solver we describe is an MHD method developed by 
\citet{Collins10} in order to study Population
III stars \citep{Xu08}, molecular clouds \citep{Collins12a}, and galaxy clusters
\citep{Xu12, Skillman13}.  A full description and suite of test problems can be
found in \citet{Collins10} and therefore we only describe the method briefly.

The divergence of the magnetic field, \Divb, is
identically zero due to the fact that the evolution of the magnetic field is the
curl of a vector, and the divergence of the curl of a vector is identically
zero.
%\begin{eqnarray}
%\partial \Bvec/\partial t & =& \frac{1}{a} \curl \Evec\\
%\Divb & =& 0.
%\end{eqnarray}
The Constrained Transport (CT) method \citep{Evans88, Balsara99} for
magnetohydrodynamic (MHD) evolution employs this same vector property
to evolve the magnetic field in a manner that preserves $\Divb=0$.  
The electric field is computed using
the fluxes from the Riemann solver.  The curl of that electric field
is then used to update the magnetic field.  The advantage of this
method is that it preserves \Divb\ to machine precision.  The primary
drawback is increased algorithm complexity.  Note also that since 
only the update of the magnetic field is divergence free, any monopoles 
created by other numerical sources persist.

The base Godunov method, described in \citet{Li08a}, uses spatially and
temporally second order reconstruction (both MUSCL-Hancock and
Piecewise Linear Method), and a selection of Riemann solvers including
HLLC and HLLD \citep{Mignone07}, as described earlier.  
The constrained transport methods
are the first-order method described by \citet{Balsara99} and the
second order methods described in \citet{Gardiner05}.  The AMR
machinery is described by \citet{Balsara01} and \citet{Collins10}.

@@

Cosmological MHD can be treated in a number of ways.  As described in 
the previous section, and equation~\ref{eq:induction} the MUSCL with Dedner
method treats the cosmological dilution of
$\vecB$ with a source term,  However, one can eliminate the right hand side of Equation 
\ref{eq:induction} by defining
$\vecB^\prime = \vecB / \sqrt{a}.$
Then equation~\ref{eq:induction} becomes
$$ 
\frac{\partial \vecB^\prime}{\partial t} - \frac{1}{a}  \curl (\vecv \times
\vecB^\prime)  = 0, \label{eq:induction_cosmo_prime_nosource}
$$
and we can reuse each part of the non-cosmological algorithm (hyperbolic solver,
electric field creation, interpolation, projection, and flux correction) without
modification.

The dual energy formalism has also been incorporated in two possible ways, one internal
energy, as described in Section \ref{sec.hydro.ppm}, and one that
uses entropy \cite{TVD93, Collins10}..  
%The additional equations are
%\begin{eqnarray}
%\end{eqnarray}
%\begin{eqnarray}
%  \frac{\partial S}{\partial t} +\frac{1}{a} \div (S \vecv) & = &  -
%  \frac{ 3 (\gamma - 1) \dot{a}}{a} S \\
%  \frac{\partial \rho e}{\partial t} + \frac{1}{a} \div (\rho e \vecv) &
%  = & -\frac{ 3(\gamma -1 ) \dot{a}}{a} + \frac{p}{a} \div \vecv,
%\end{eqnarray}
%where $e$ is the internal energy and $S=p/\rho^{\gamma-1}$ is the entropy.  If
%the ratio of thermal to kinetic energy, $\eta$, is below a certain value, the
%alternative energy is used.  As in \citet{Li08a}, we use $\eta=0.008$.

%
%\enzo's MHD-CT has been used in a variety of contexts, from galaxy
%cluster evolution \citep{Xu11,Xu12}, synthetic observations of radio
%halos \citep{Skillman13}, primordial star formation \citep{Xu08}, and
%present-day star formation \citep{Collins11, Collins12a}.
% GB NOTE: incorporate these elsewhere

### 2013-06-08 numerical-mhd-dedner.tex
Finished up my edits on the fluid equations.

The second method we describe is a MUSCL-based solver than can be used in both HD and MHD modes.  To more clearly include the Dedner divergence cleaning modifications, we write the equations of compressible inviscid hydrodynamics can be written in the form of conservation laws as,
\begin{equation}
 \frac{\partial{U}}{\partial{t}} +
 \frac{\partial{F^x}}{\partial{x}} + \frac{\partial{F^y}}{\partial{y}} + \frac{\partial{F^z}}{\partial{z}}= 0, \label{hydro}
\end{equation}
The conserved variable $U$ is given by
\begin{equation}
 U = (\rho, \rho v_x, \rho v_y, \rho v_z, \rho E)^{T},
\end{equation} 
where $\rho$ is density, $v_i$ are the three components of velocity
for $i={x,y,z}$, $E=v^2/2 + e$ denotes the specific total energy and $e$ the
specific internal energy.

@@

Much like the PPM solver, we have three basic steps: the first is
reconstruction of the variables, the second is a solution of the
Riemann problem, and the third is is updating the conserved quanties
with the fluxes as written above.

For the reconstruction scheme we here implemented only the simple
piecewise linear reconstruction \citep{
And a number of approximate Riemann
solvers.

The method is dimensionally un-split in that the fluxes are computed
for all dimensions first and the conserved quantities are updated in
one go in contrast to the Strang splitting employed in the Zeus and
PPM solvers described above.

>>>>

The second method we describe is a MUSCL-based solver than can be used in both HD and MHD modes.  The description here will be very brief both because the ideas are similar to those described in the previous section, and because this implementation has previously
been described in more detail elsewhere \citep{WangAbelZhang08, WangAbel09}.

Much like the PPM solver, we have three basic steps: the first is
reconstruction of the variables, the second is a solution of the
Riemann problem, and the third is is updating the conserved quanties
with the fluxes as written above.
For the reconstruction scheme we have implemented only the simple
piecewise linear reconstruction \citep{1979JCoPh..32..101V, 1985JCoPh..59..264C},
with options for both primitive and conservative variable reconstruction.
The available Riemann solvers are HLL, HLLC, and HLLD, as described earlier.

To more clearly describe the Dedner divergence cleaning modifications, we write the equations of compressible inviscid hydrodynamics can be written in the form of conservation laws as,
\begin{equation}
 \frac{\partial{U}}{\partial{t}} +
 \frac{\partial{F^x}}{\partial{x}} + \frac{\partial{F^y}}{\partial{y}} + \frac{\partial{F^z}}{\partial{z}}= 0, \label{hydro}
\end{equation}
The conserved variable $U$ is given by
\begin{equation}
 U = (\rho, \rho v_x, \rho v_y, \rho v_z, \rho E)^{T},
\end{equation} 
where $\rho$ is density, $v_i$ are the three components of velocity
for $i={x,y,z}$, $E=v^2/2 + e$ denotes the specific total energy and $e$ the
specific internal energy.

@@

The method is dimensionally un-split in that the fluxes are computed
for all dimensions first and the conserved quantities are updated in
one step, in contrast to the Strang splitting employed in the other solvers.
Also unlike the other schemes, time-integration is done with a second-order
Runge-Kutta scheme \citep{1988JCoPh..77..439S}.

### 2013-06-08 numerical-zeus.tex
Finished up my edits on the fluid equations.

As an alternative to PPM, \enzo\ also includes an implementation of the finite-difference hydrodynamic algorithm employed in the compressible magnetohydrodynamics code \zeus\ \citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} within a multistep (operator split) solution procedure that is fully explicit in time.  This method is formally second order-accurate in space but first order-accurate in time.

@@

The \zeus\ method uses a von Neumann-Richtmeyer artificial viscosity to smooth shock discontinuities that may appear in fluid flows and can cause a breakdown of the finite-difference equations.  The artificial viscosity term is added in the source terms as:

\begin{eqnarray}
\rho \frac{\partial\vecv}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \vecv - \textbf{Q} : \grad \vecv, 
\end{eqnarray}
%where \textbf{v} is the baryon velocity, $\rho$ is the mass density, $p$ is the pressure, $e$ is the internal energy density of the gas and

Here \textbf{Q} is the artificial viscosity stress tensor, which we take to be diagonal with on-axis terms given by $l^2 \rho (\partial v / \partial x)^2$ as proposed by von Neumann \& Richtmyer.  The length scale $l$ determines the width of shocks and is typically a few times the cell spacing.

@@

As in the original \zeus\ paper, the source terms are added in three steps. First we add the pressure and gravity forces:
\begin{equation}
v_j^{n+a}  =  v_j^n - \frac{\Delta t}{\Delta x_j} \frac{p^n_j - p^n_{j-1}} {(\rho^n_j + \rho^n_{j-1})/2}.
\end{equation}
Partial updates are denoted by the $n+a$, $n+b$ notation.  We show updates in one dimension as the extension to the multi-dimensional case is straightforward (note that all dimensions are carried out for each substep before progressing to the next substep). We then add the artificial viscosity:

\begin{eqnarray}
v_j^{n+b} & = & v_j^{n+a} - \frac{\Delta t}{\Delta x_j} 
                             \frac{q^{n+a}_j - q^{n+a}_{j-1}} {(\rho^n_j + \rho^n_{j-1})/2} \\
e_j^{n+b} & = & e_j^n - \frac{\Delta t}{\Delta x_j} q^{n+a}_j (v^{n+a}_{j+1} - v^{n+a}_{j}).
\end{eqnarray}

The artificial viscosity coefficient $q_j$ is given by:
\begin{equation}
q_j = \left\{ \begin{array}{ll}
              Q_{\rm AV} \rho_j (v_{j+1} - v_j)^2 \quad & \rm{if} (v_{j+1} - v_j) < 0 \\
               0 & \rm{otherwise}
               \end{array} \right.
\end{equation}
where $Q_{\rm AV}$ is a constant with a typical value of 2. We refer the interested reader to \citet{Stone92a} and \citet{1994ApJ...429..434A} for more details.  We also include the option (turned off by default) of adding a linear artificial viscosity as suggested in the \zeus\ paper for stagnant flow regions.  This is given by
\begin{equation}
q_{{\rm lin},j} = Q_{\rm LIN} \rho c_j (v_{j+1} - v_{j})
\end{equation}
where $c_j^2 = \gamma p/\rho$ is the adiabatic sounds speed.

@@

A limitation of a technique that uses an artificial viscosity is that, while the correct Rankine-Hugoniot jump conditions are achieved, shocks are broadened over 3-4 mesh cells and are thus not treated as true discontinuities. This may cause unphysical pre-heating of gas upstream of the shock wave, as discussed in \citet{1994ApJ...429..434A}.  On the other hand, it is much more robust than PPM and is easy to add additional physics.  We also note that this method solves only the internal energy equation rather than total energy, so the ``dual energy formulation'' discussed in Section~\ref{sec.hydro.ppm} is unnecessary.

>>>>

As an alternative to the previous Godunov methods, \enzo\ also includes an implementation of the finite-difference hydrodynamic algorithm employed in the compressible magnetohydrodynamics code \zeus\ \citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} within a multistep (operator split) solution procedure that is fully explicit in time.  This method is formally second order-accurate in space but first order-accurate in time.

@@

The \zeus\ method uses a von Neumann-Richtmeyer artificial viscosity to smooth shock discontinuities that may appear in fluid flows and can cause a breakdown of the finite-difference equations.  The artificial viscosity term is added in the source terms as:
\begin{eqnarray}
\rho \frac{\partial\vecv}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \vecv - \textbf{Q} : \grad \vecv, 
\end{eqnarray}
Here \textbf{Q} is the artificial viscosity stress tensor, which we take to be diagonal with on-axis terms given by $l^2 \rho (\partial v / \partial x)^2$ as proposed by von Neumann \& Richtmyer.  The length scale $l$ determines the width of shocks and is typically a few times the cell spacing.

@@

As in the original \zeus\ paper, the source terms are added in three steps. First we add the pressure and gravity forces:
\begin{equation}
v_j^{n+a}  =  v_j^n - \frac{\Delta t}{\Delta x_j} \frac{p^n_j - p^n_{j-1}} {(\rho^n_j + \rho^n_{j-1})/2}.
\end{equation}
Partial updates are denoted by the $n+a$, $n+b$ notation.  We show updates in one dimension as the extension to the multi-dimensional case is straightforward (note that all dimensions are carried out for each substep before progressing to the next substep). We then add the artificial viscosity:
\begin{eqnarray}
v_j^{n+b} & = & v_j^{n+a} - \frac{\Delta t}{\Delta x_j} 
                             \frac{q^{n+a}_j - q^{n+a}_{j-1}} {(\rho^n_j + \rho^n_{j-1})/2} \\
e_j^{n+b} & = & e_j^n - \frac{\Delta t}{\Delta x_j} q^{n+a}_j (v^{n+a}_{j+1} - v^{n+a}_{j}).
\end{eqnarray}
The artificial viscosity coefficient $q_j$ is given by:
\begin{equation}
q_j = \left\{ \begin{array}{ll}
              Q_{\rm AV} \rho_j (v_{j+1} - v_j)^2 \quad & \rm{if} (v_{j+1} - v_j) < 0 \\
               0 & \rm{otherwise}
               \end{array} \right.
\end{equation}
where $Q_{\rm AV}$ is a constant with a typical value of 2. We refer the interested reader to \citet{Stone92a} and \citet{1994ApJ...429..434A} for more details.  We also include the option (turned off by default) of adding a linear artificial viscosity as suggested in the \zeus\ paper for stagnant flow regions.  This is given by
\begin{equation}
q_{{\rm lin},j} = Q_{\rm LIN} \rho c_j (v_{j+1} - v_{j})
\end{equation}
where $c_j^2 = \gamma p/\rho$ is the adiabatic sounds speed.

@@

A limitation of a technique that uses an artificial viscosity is that, while the correct Rankine-Hugoniot jump conditions are achieved, shocks are broadened over 3-4 mesh cells and are thus not treated as true discontinuities. This may cause unphysical pre-heating of gas upstream of the shock wave, as discussed in \citet{1994ApJ...429..434A}.  On the other hand, it is much more robust than PPM and is easy to add additional physics.  We also note that this method solves only the internal energy equation rather than total energy, so the `dual energy formulation' discussed in Section~\ref{sec.hydro.ppm} is unnecessary.
